Adaptive evolution with aneuploidy and mutation

### 2021-10-20 ms/ms.tex
small todo was added about priors of the model without aneuploidy
added likelihood function for no aneuploidy model, updated likelihood figure

\paragraph{Likelihood function.}
Denote by $A_{t}$ the fixation of aneuploidy at generation $t$, by $L_{t}$ the loss of aneuploidy at generation $t$, and by $L^*_{t,T}$ the loss of aneuploidy at generation $t$ conditioned on no loss by generation $T$.
The model likelihood function for parameter vector $\theta$ for the heat-stress experiment is
\begin{multline} \label{eq:heatstress-likelihood}
\likelihood(\theta | A_{450}, L_{1700}, L^*_{2350, 1700} ) = \\
P^4(A_{450}) \cdot \Big(1 - P^4(\lnot{L_{1700}} | A_{450} ) - P^4(\lnot{L^*_{2350, 1700}}| A_{450}) + P^4(\lnot{L_{1700}} \land \lnot{L^*_{2350, 1700}}| A_{450})\Big)	\;,	
\end{multline}
where $P(X)$ is the probability of event $X$ and $\lnot$ means \emph{not}.
The likelihood function of the model with parameter set $\theta$ for the pH-stress experiment is
\begin{equation} \label{eq:phstress-likelihood}
\likelihood(\theta | A_{150}, L_{750} ) = \\
 P^4(L_{450}) \cdot 6 \cdot	P^4(L_{750} | A_{150}) \cdot P^4(\lnot{L_{750}} | A_{150}) \;.  %%TODO i think it is wrong because fig 5C shows that only 2 replicas are fixated after 650 generations (see Supplementary also)
\end{equation} 
Because of the complexity of the model, this likelihood function is intractable. Therefore, given specific values of the model parameters $\theta$, we simulate the model in many replicates to approximate the value of the likelihood function $\likelihood(\theta)$.

@@

\paragraph{Prior distributions.} The prior distributions of $w_{2n+1}=1-c+b$, $w_{2n+1^*}=(1+s)(1-c)+b$ and $w_{2n^*}=1+s$ were obtained by estimating $1-c$, $1-c+b$ and $1+s$ from growth curves data. These growth curve data were previously obtained by~\citet{Yona2012} using mono-culture growth experiments.
The raw growth curves data were not published before, but were used to produce figures 3C, 4A, and S2 in \citet{Yona2012}.
We used \texttt{Curveball}\footnote{https://curveball.yoavram.com}, a dedicated method for predicting results of competition experiments from growth curve data~\citep{Ram2019}. \texttt{Curveball} takes growth curves of two strains growing separately in mono-culture and predicts how they would grow in a mixed culture, that is, it predicts the results of a competition assay.
From these predictions, relative fitness values can be computed.  
We used growth curves of $2n+1$ and $2n$ in \SI{39}{\celsius} and in \SI{30}{\celsius} to estimate $1-c+b$ and $1-c$, respectively. The latter assumes that the cost of aneuploidy is the same in \SI{39}{\celsius} and \SI{30}{\celsius}; this might be incorrect, but we only assume this to generate a prior distribution for the fitness values. We also used growth curves of $2n+1$ and $2n^*$ in \SI{39}{\celsius} to estimate $1+s$. 
Because \texttt{Curveball} uses a maximum-likelihood approach to estimate model parameters, we were able to estimate a distribution of relative fitness values by sampling from a truncated multivariate normal distribution defined by the maximum-likelihood covariance matrix. Thus, we sampled 10,000 values for $1-c+b$, $1-c$, and $1+s$, which we used as prior distributions. Note that the first and last values correspond to $w_{2n+1}$ and $w_{2n}$, whereas $w_{2n+1^*}=(1-c)(w_{2n^*}-1)+w_{2n+1}$.
See Figures~\ref{fig:growth-curves-30deg} and~\ref{fig:growth-curves-39deg}.
%TODO ILIA please report here the point estimate and 95% CI for the three distributions, I only have for 1-c+b: 0.9863 (95% CI 0.9790 - 0.9939).
%TODO ILIA describe the priors for the mutation and aneuploidy rate.

@@

% likelihood plots
\begin{figure}[b!]
  \centering
  \begin{subfigure}[t]{0.5\textwidth}
      \caption{
        \textbf{A. With aneuploidy}
      }
      \centering
      \includegraphics[height=1.8in]{../figures/likelihood-plot.pdf}      
      \label{fig:likelihood-with-aneuploidy}
  \end{subfigure}%
  \\
  \begin{subfigure}[t]{0.5\textwidth}
  	  \caption{
        \textbf{B. Without aneuploidy}
      }
      \centering
      \includegraphics[height=1.8in]{../figures/likelihood-plot.pdf}      
      \label{fig:likelihood-without-aneuploidy}
  \end{subfigure}
  \caption{
    \textbf{Single-locus model fit.} Cumulative distribution of 10000 simulations are shown.
    \textbf{(A)} Simulations parameters are from the median of the posterior of the single-locus model fit. In close to 50\% of simulations \eumt fixation occurs between 450 and 1700 generations and in about 50\% \eumt fixation occurs between 1700 and 2350 generatyions. This results in high likelihood.
    \textbf{(B)} Simulations parameters are from the median of the posterior of the single-locus model \emph{without} aneuploidy fit. TODO
  }
  \label{fig:likelihood}
\end{figure}

>>>>

\paragraph{Likelihood function.}
Denote by $A_{t}$ the fixation of aneuploidy at generation $t$, by $L_{t}$ the loss of aneuploidy at generation $t$, and by $L^*_{t,T}$ the loss of aneuploidy at generation $t$ conditioned on no loss by generation $T$.
The model likelihood function for parameter vector $\theta$ for the heat-stress experiment is
\begin{multline} \label{eq:heatstress-likelihood}
\likelihood(\theta | A_{450}, L_{1700}, L^*_{2350, 1700} ) = \\
P^4(A_{450}) \cdot \Big(1 - P^4(\lnot{L_{1700}} | A_{450} ) - P^4(\lnot{L^*_{2350, 1700}}| A_{450}) + P^4(\lnot{L_{1700}} \land \lnot{L^*_{2350, 1700}}| A_{450})\Big)	\;,	
\end{multline}
where $P(X)$ is the probability of event $X$ and $\lnot$ means \emph{not}.

The likelihood function of the model with parameter set $\theta$ for the pH-stress experiment is
\begin{equation} \label{eq:phstress-likelihood}
\likelihood(\theta | A_{150}, L_{750} ) = \\
 P^4(L_{450}) \cdot 6 \cdot	P^4(L_{750} | A_{150}) \cdot P^4(\lnot{L_{750}} | A_{150}) \;.  
 \end{equation} 
%TODO i think it is wrong because fig 5C shows that only 2 replicas are fixated after 650 generations (see Supplementary also) Anyway, I don't think that we use this likelihood, so probably should be removed at all.

For the model without aneuploidy let's denote by $M_{t}$ the fixation of \eumt at generation $t$, and by $M^*_{t,T}$ the fixation of \eumt at generation $t$ conditioned on no fixation at generation $T$.
The likelihood of the model without aneuploidy with parameter vector $\theta$ for the heat-stress experiment is
\begin{multline} \label{eq:heatstress-noaneuploidy-likelihood}
\likelihood(\theta |  M_{1700}, M^*_{2350, 1700} ) = \\
\Big(1 - P^4(\lnot{M_{1700}}) - P^4(\lnot{M^*_{2350, 1700}}) + P^4(\lnot{M_{1700}} \land \lnot{M^*_{2350, 1700}})\Big)	\;,	
\end{multline}

Because of the complexity of the model, this likelihood function is intractable. Therefore, given specific values of the model parameters $\theta$, we simulate the model in many replicates to approximate the value of the likelihood function $\likelihood(\theta)$.

@@

\paragraph{Prior distributions.} The prior distributions of $w_{2n+1}=1-c+b$, $w_{2n+1^*}=(1+s)(1-c)+b$ and $w_{2n^*}=1+s$ were obtained by estimating $1-c$, $1-c+b$ and $1+s$ from growth curves data. These growth curve data were previously obtained by~\citet{Yona2012} using mono-culture growth experiments.
The raw growth curves data were not published before, but were used to produce figures 3C, 4A, and S2 in \citet{Yona2012}.
We used \texttt{Curveball}\footnote{https://curveball.yoavram.com}, a dedicated method for predicting results of competition experiments from growth curve data~\citep{Ram2019}. \texttt{Curveball} takes growth curves of two strains growing separately in mono-culture and predicts how they would grow in a mixed culture, that is, it predicts the results of a competition assay.
From these predictions, relative fitness values can be computed.  
We used growth curves of $2n+1$ and $2n$ in \SI{39}{\celsius} and in \SI{30}{\celsius} to estimate $1-c+b$ and $1-c$, respectively. The latter assumes that the cost of aneuploidy is the same in \SI{39}{\celsius} and \SI{30}{\celsius}; this might be incorrect, but we only assume this to generate a prior distribution for the fitness values. We also used growth curves of $2n+1$ and $2n^*$ in \SI{39}{\celsius} to estimate $1+s$. 
Because \texttt{Curveball} uses a maximum-likelihood approach to estimate model parameters, we were able to estimate a distribution of relative fitness values by sampling from a truncated multivariate normal distribution defined by the maximum-likelihood covariance matrix. Thus, we sampled 10,000 values for $1-c+b$, $1-c$, and $1+s$, which we used as prior distributions. Note that the first and last values correspond to $w_{2n+1}$ and $w_{2n}$, whereas $w_{2n+1^*}=(1-c)(w_{2n^*}-1)+w_{2n+1}$.
See Figures~\ref{fig:growth-curves-30deg} and~\ref{fig:growth-curves-39deg}.
%TODO ILIA please report here the point estimate and 95% CI for the three distributions, I only have for 1-c+b: 0.9863 (95% CI 0.9790 - 0.9939).
%TODO ILIA describe the priors for the mutation and aneuploidy rate.
% TODO also describe the prior for the model without aneuploidy - the mutation rate is wider here.

@@

% likelihood plots
\begin{figure}[b!]
  \centering
  \begin{subfigure}[t]{0.5\textwidth}
      \caption{
        \textbf{A. With aneuploidy}
      }
      \centering
      \includegraphics[height=1.8in]{../figures/likelihood-plot.pdf}      
      \label{fig:likelihood-with-aneuploidy}
  \end{subfigure}%
  \\
  \begin{subfigure}[t]{0.5\textwidth}
  	  \caption{
        \textbf{B. Without aneuploidy}
      }
      \centering
      \includegraphics[height=1.8in]{../figures/likelihood-noaneuploidy-plot.pdf}      
      \label{fig:likelihood-without-aneuploidy}
  \end{subfigure}
  \caption{
    \textbf{Single-locus model fit.} Cumulative distribution of 10000 simulations are shown.
    \textbf{(A)} Simulations parameters are from the median of the posterior of the single-locus model fit. The likelihood for the heat-stress experiment (\autoref{eq:heatstress-likelihood}) is 0.86.
    \textbf{(B)} Simulations parameters are from the median of the posterior of the single-locus model \emph{without} aneuploidy fit. The likelihood for the heat-stress experiment (\autoref{eq:heatstress-noaneuploidy-likelihood}) is 0.75.
  }
  \label{fig:likelihood}
\end{figure}
